---
title: "Analysis of colours and learning"
author: "Timothée Flutre"
date: "Friday, August 29, 2014"
output: html_document
---

License: Creative Commons BY-NC-SA

The program "Nouveaux Commanditaires Science" aims at ...
It is organized by the Atelier des Jours à Venir and funded by the Fondation de France.

This document was initially developped for the group in Mollins, but can be used as a template for any group willing to investigate a research question.
Have fun!


## Define the question

TODO


## Design the experiment

TODO: the protocol is ...


## Simulate some data

Before having acquired real data, we can still start working by generating fake data:
```{r simul.data}
set.seed(2014)
nb.words = 36
nb.people = 50
p1 = 0.3
p2 = 0.4
y1 = rbinom(n=nb.people, size=nb.words, prob=p1)
y2 = rbinom(n=nb.people, size=nb.words, prob=p2)
```


## Load, format and clean the real data

TODO


## Explore the data visually

Let´s plot the raw data:
```{r plot}
tmp = cbind(y1, y2)
colnames(tmp) = c("white", "colour")
plot(x=jitter(rep(1,nb.people)), y=y1, xlim=c(0,3), ylim=c(0, nb.words),
     xaxt="n", xlab="", ylab="nb of memorized words", las=1,
     main=paste0("persons=",nb.people, "   p1=", p1, "   p2=", p2))
axis(side=1, at=c(1,2), labels=c("white", "colour"))
points(x=jitter(rep(2, nb.people)), y=y2)
```

TODO: use segments() to show that data are paired

We can also make two boxplots (especially if there are many samples):
```{r boxplot}
boxplot(tmp, main=paste0("persons=",nb.people, "   p1=", p1, "   p2=", p2),
        ylim=c(0,nb.words), ylab="nb of memorized words", las=1, notch=TRUE)
```


## Write the model

Likelihood: $y_{1i} \sim \mathcal{B}(N, p_1)$ and $y_{2i} \sim \mathcal{B}(N, p_2)$

Is $p_1 = p_2$?


## Fit the model

Test the null hypothesis "no effect of background color on memory":

```{r test}
t.test(y1, y2, alternative="two.sided", paired=TRUE)
```


## And now?

Check the model, then change it? Get more data? Share the results? ...
